---
title: "<i class='fas fa-pencil-alt' style='margin-left:4px;margin-right:4px;'></i> Precourse"
output:
  bookdown::html_document2:
    toc: false
    number_sections: false
---

```{r,child="assets/header-lab.Rmd"}
```

```{r,eval=TRUE,include=FALSE}
library(yaml)
upid <- yaml::read_yaml("_site.yml")$uppmax_project
```

These are steps to be completed before the workshop.

* A remote computing cluster (UPPMAX) will be use for handling raw sequence data. A SUPR/SNIC account is needed to use UPPMAX resources. If you do not already have one, [create an account](https://supr.snic.se/). Log in to [SUPR/SNIC](https://supr.snic.se/) and request membership to the project ID **`r paste0(upid)`**. 

* Make sure you can connect to the Uppmax cluster Rackham through a terminal. Type `ssh -Y user@rackham.uppmax.uu.se` and then enter your password. The password will not be visible as you type. 

<div class="instruction">
<i class="fas fa-lightbulb"></i> If you are on a Windows system, we recommend [MobaXterm](http://mobaxterm.mobatek.net). For MobaXterm users, it is recommended that you install the program and not use the portable version. MobaXterm also has an integrated SFTP file browser.  
<i class="fas fa-lightbulb"></i> For all users, if you prefer a GUI to upload and download files from UPPMAX, we recommend installing [FileZilla](https://filezilla-project.org/).
</div>

* If you do not feel comfortable working in a terminal, please take the time to do the first three parts in the *[Unix tutorial for beginners](http://www.ee.surrey.ac.uk/Teaching/Unix/)* before the workshop. Otherwise, it may be difficult for you to follow the exercises during the workshop. 

* Make sure that you can read and write in the workshop folder. Go to ``r paste0("/proj/",upid,"/nobackup/")`` and create a directory with your username. For example `mkdir jody`. You will work inside this directory for the workshop. For example ``r paste0("/proj/",upid,"/nobackup/jody")``. If you cannot write to the folder, the most likely reason is that you have not requested access to the workshop project via SUPR.  

<div class="instruction">
<i class="fas fa-exclamation-circle"></i> Note that it may take an hour or so from request approval, before you can actually write to the folder. We will check before the workshop that all students have logged in and done this, so do not forget!  
</div>

* Download IGV on your own computer and make sure that you have hg19 genome available. The data we will work with is from human and we will visualise the results in the IGV genome browser. IGV can be downloaded from the [Broad Institute](https://www.broadinstitute.org/igv/).

* Download and install the latest version of R from [r-project.org](https://cloud.r-project.org/). If you have an old version of R and you do not use it, uninstall it and then install a newer version. Make sure you have one of the recent versions of R. For Windows users, it is recommended that you DO NOT install to `C:\Program Files\R\`. Instead, install to `C:\R\`. 

* Install [RStudio](https://www.rstudio.com/products/rstudio/download/). RStudio provides you with tools like code editor with highlighting, project management, version control, package building, debugger, profiler and more.

* Extra R packages used in the workshop exercises (if any) are listed below. It is recommended that you install this in advance. Simply copy and paste the code into R.

```{r,eval=TRUE,block.title=NULL,echo=FALSE,comment="",class.output="r"}
# this code block reads package names from '_site.yml' and prints them as installation instruction.

pkg <- yaml::read_yaml("_site.yml")

if(!is.null(pkg$packages$packages_cran_student)) {
 cat("# install from cran\n")
 cat(paste0("install.packages(c('",paste(pkg$packages$packages_cran_student,sep="",collapse="','"),"'))"))
 cat("\n")
}

if(!is.null(pkg$packages$packages_bioc_student)) {
 cat("# install from bioconductor\n")
 cat(paste0("BiocManager::install(c('",paste(pkg$packages$packages_bioc_student,sep="",collapse="','"),"'))"))
}
```

**End of document.**
