---
title: "Online Repositories"
subtitle: "Workshop on RNA-Seq"
editor_options:
  chunk_output_type: console
---

```{r,child="assets/header-lab.Rmd"}
```

<div class="boxy boxy-exclamation boxy-yellow">
Create a directory named `data` in your current working directory for input and output files.
</div>

Load R packages.


```{r, eval=F}
# BiocManager::install("GEOquery")
library(GEOquery)
library(Biobase)
library(rafalib)
rafalib::mypar(mar=c(6,2.5,2.5,1)) #sets nice arrangement for the whole document

# source download function
source("https://raw.githubusercontent.com/NBISweden/workshop-RNAseq/master/assets/scripts.R")
```


## Loading a dataset from GEO using GEOquery

For this exercise, let's try getting the full count matrix from the original  [GSE131032](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE131032) dataset.

In some cases the simple `getGEO` will download the data, metadata and gene annotation into a ExpressionSet format, where you can then extract the information about the gene counts, metadata and gene annotations. 

```{r, eval=F}
#Get the GSE object from it
gset <- GEOquery::getGEO("GSE131032")[[1]]
gset

#Get the count matrix from the GSE objetc
data <- gset@assayData$exprs
dim(data)

#Get the sample metadata / phenotypic from the GSE object 
metadata <- gset@phenoData@data
dim(metadata)

#Get the gene annotation data from the GSE object 
annot <- gset@featureData@data
dim(annot)
```

However, in most cases (in our experience) that is not the case since every dataset is a bit unique. So we recommend checking the files deposited individually and download what is most relevant for you. We can first list the files deposited: 

```{r, eval=F}
file_list <- GEOquery::getGEOSuppFiles("GSE131032",fetch_files = F)
print(file_list)
```

You can see the `RAW.tar` file, which contains the original count file deposited individually for every sample (under the GSM). The command below will dowlaod the file into a folder with the GSE name. This is a `.tar` compressed file that can be extracted.

```{r, eval=F}
GEOquery::getGEOSuppFiles("GSE131032",fetch_files = T,filter_regex = "_RAW.tar")
list.files("GSE131032")

#extract tar using bash (within R, but can be done outside too)
system("cd GSE131032; tar -xvf GSE131032_RAW.tar")
system("cd GSE131032; rm GSE131032_RAW.tar")

list.files("GSE131032")

system("cd GSE131032; gunzip *.gz")
```

Having the individual files, we can combine them using a for loop or apply function:

```{r, eval=F}
#list the files
list.files("GSE131032")

#Check the column names from the 1st file
head( read.delim(paste0("GSE131032/",list.files("GSE131032")[1])) )

#Do an apply function to get all the estimated counts from each file and compile into a single matrix
data <- sapply(list.files("GSE131032",pattern = ".tsv"), function(x){
  x <- read.delim(paste0("GSE131032/",x))[,"est_counts"]
  return(x)
})
rownames(data)
colnames(data)

#Add the gene names and sample names
rownames(data) <- read.delim(paste0("GSE131032/",list.files("GSE131032")[1]))[,"target_id"]
colnames(data) <- sub("_.*","",colnames(data))

#Check the column names from the 1st file
rownames(metadata) == colnames(data)

#Rename data
colnames(data) <- metadata$title
head(data)
```

Alternativelly, in the `file_list` above we could also simply load the count matrix.

```{r, eval=F}
#list files available
print(file_list)

#Download  and extract the full matrix
GEOquery::getGEOSuppFiles("GSE131032",fetch_files = T,filter_regex = "GSE131032_kallisto_counts.csv.gz")
system("cd GSE131032; gunzip GSE131032_kallisto_counts.csv.gz")

#Read the file (note that the sample names here are not the GSM accesccion, but the investigator's IDs)
data <- read.delim(paste0("GSE131032/GSE131032_kallisto_counts.csv"),sep = ",",row.names = 1)
colnames(data)

#Those sample IDs are stored in one of the metadata columns
metadata$description
metadata$description == colnames(data)

#Rename data
colnames(data) <- metadata$title
head(data)
```

